What We Eval in the Shadows

### 2021-04-13 paper/main.tex
envirnments

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}
% Other macros that can be used:
% - \packageMinimizedMedianOperations: median number of operations
% - \packageMinimizedPercentEnvir: percentages of sites with the default argument
% - \packageMinimizedPercentCallsEnvir: percentage of calls with the default envir argument

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically aren't,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
\eval. For the other cases, either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps, in fact the median is only
\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
mostly irrelevant (unless it redefines a built-in operator).
%% Why then this variety of environments??

\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case.

\newpage
\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
occurs in 6\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations.

@@

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls.Assignments are excluded from this form. They
occurs in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd.

@@

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statements
code blocks, which occur in only 4\% of sites. These are larger expressions, we
do not inspect contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd.

\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
returns a dataframe that is the result of fitting the model described in a given
formula. This form subsumes \c{F(F(X))}, \c{FUN} and assignments. It is the
single most popular function invoked from \eval, it occurs in 7\% of the sites.

@@

The environment in which evaluation happens determines what is visible to the
computation started by \eval, and the potential reach of its side-effects. The
second argument to \eval, \c{envir}, defines that environment. It can be
omitted, in which case its defaults to \c{parent.frame()} which returns the
caller's environment. Other possible values for \c{envir} include, lists,
dataframes, pairlists, integers and \c{Null}. Lists and dataframe are coerced to
new environments, integer values are passed to \c{sys.call} which returns the
environment of a stack frame at the requested offset.

The top-level environment in R is called the global environment. New
environments can be created using \c{new.env}, or \c{list2env}. They can be
provided a parent environment which becomes the enclosing scope of the new
environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
% as.environment also can be used sometimes for that
% list2env creates an environment only if its argument envir is not provided

Another kind of environments that is passed to \eval is the environment of a
package or its \emph{namespace}. A call to
\c{library(packageName)}, which loads libraries in R, for instance, will
populate environment \c{package:packageName}. It is a \emph{base} function
though, so we do not see it in the CRAN dataset. However, R still makes it
possible to access any package environments and \packageNbPackageEnvSites call
sites, \ie \packagePackageEnvSitePercent access a package environment, from
\packageNbPackageEnvPackages different packages. Those packages perform
introspection, to represent and search any environment by name, in package
\emph{envnames}, to containerize tests, as with \emph{testthat} and
\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
poor man's hidden functions in \emph{USSCXenaTools}. It is also used to
implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.

%TODO: there is a difference between a package environment and a namespace. Shortly explain.

Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
passed an actual environment.

It can also be \c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
list or data frame and creates bindings for them in a new environment.

We classify environments to \eval as follows:

\begin{itemize}
	\item function environments: the frame of the calling function or one of the parent frames (obtained using \c{parent.frame()}, \c{sys.frame()} or with an integer argument for \c{envir} for instance);
	\item list environments: a list, a data frame, a tibble, or \c{NULL} (seen as the empty list), passed to \eval;
	\item the global environment;
	\item package and namespace environments;
	\item the empty environment (created with \c{emptyenv()})
    \item new environments, created with \c{new.env}, \c{list2env} or \c{as.environment}.
\end{itemize}

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a function environment, with \c{global} as a distant second.

\begin{table}[h]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Environment & \#sites & \%sites \\
		\hline
		Function environments &  \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
		Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
		List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
		Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\
		Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
		\hline
		New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
		\hline
	\end{tabular} \label{tab:highlevelenvironments}
	\caption{Classification of environments per site.}
\end{table}

Table~\ref{tab:framenumber} shows the function environments classified according to their position in the call stack: $0$ is the current function, $1$ the parent function and so on.

The calls happening in a function environment are mostly happening in the environment of the function calling \eval. This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument. A lot of sites also use \c{parent.frame()} to access their parent frame. It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level. Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites. Accessing further-away frames happens much more seldom.

\begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        Frame number & \#sites & \%sites \\
        \hline
        \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
        \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
        \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
        > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\
        \hline
    \end{tabular} \label{tab:framenumber}
\caption{Frame number for the function environments. One site can possibly be passed different kinds of environments.}
\end{table}

A new environment creation, a parent environment must be specified. In Table~\ref{tab:newenvs}, we show what environments the new environment inherits from ultimately.\footnote{It means we directly show the first parent that is different from a new environment.} New environments mostly inherits from function environment: that is actually the default for \c{new.env}, with argument \c{parent = parent.frame()}. A large number of new environments are also wrapping package environment or namespaces. This is a widespread pattern to store variables to use at the package level. When a new environment wraps the empty environment, it can be handily used as a hash table.\footnote{\c{new.env} builds an environment that stores the variable name with a hash table by default. }
% Speak also about new and global?

\begin{table}[h]
    \centering 
    \begin{tabular}{|c|c|c|}
        \hline
        Category & \#sites & \%sites \\
        \hline
        \packageNewEnvCategoryNamea & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
        \packageNewEnvCategoryNameb & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
        \packageNewEnvCategoryNamec & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
        \packageNewEnvCategoryNamed & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\
        \hline
    \end{tabular} \label{tab:newenvs}
\caption{Classification of environments wrapped by one or several \c{new}, in number of sites. Some sites can be passed more than one kind of environments. }
\end{table}

% Up-to-date
% Probably to remove
\begin{table}[htbp]{ \centering
        \begin{tabular}{|c|c|}\hline
            Environment Class & Sites\\\hline
            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
            \hline
    \end{tabular}} \label{tab:environments}
    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
\end{table}

\subsubsection{Bits of text, possibly reusable}

This pattern is used to evaluate formulas which can directly refer to the fields
of the data.

We observe that a disproportionately high number of core R \eval calls access
the caller's environment. This is because many R functions call functions that
pass the result of \c{parent.frame()} to an \eval. A disproportionately high
number of calls to \eval happens in an extended top-level environment. This can
be explained by the fact that many packages evaluate code passed from the user's
workspace in the top-level environment to access the bindings.

The \c{imchange} function of package \c{imager} makes it possible to modify
images using a dedicated formula syntax.\footnote{Inspired by {map} in package
\emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a new
environment that inherits from \c{parent.frame()} by default (classified as 1+).

\begin{lstlisting}
 newenv <- new.env()
 ...
 fo <- parse(text=as.character(fo)[2])
 im[where] <- eval(fo,envir=newenv,enclos=env)
\end{lstlisting}

\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a
function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
transformed into an auxiliary function \c{h2} that can be optimized. Here, the
list used for \c{envir} ensures the correspondance between the textual arguments
of \c{h} and the arguments of \c{h2}.

\begin{lstlisting}
 sh <- substitute(h)
 fcall <- paste(sh, "(x, y)")
 ...
 h2 <- function(x, y)
 eval(parse(text = fcall),
 envir = list(x = x, y = y),
 enclos = parent.frame(2))
\end{lstlisting}

\paragraph{\c{enclos} argument.} The \c{enclos} argument of \eval can be provided when \c{envir} is a pairlist or a dataframe. It indicates in which environment to look for elements not present in \c{envir}. In our dataset, \c{enclos} is passed explicitly to \eval in \packageEnclosSitePercent of the sites.
% 85% (\packageEnclosWithListPercent) actually have a list \c{(envir)}. The rest of
% of them have environment \c{envir}. In that case, \eval ignores \c{enclos} anyway

\todo{That should be moved to the side effect section and updated with the last data.}
Categorize side-effects between env passed to eval and other envs:
\begin{itemize}
\item ephemeral -- read to envs that are created during \eval  XXX033
\item local -- reads to env of the function in which the call to \eval
  lexically occurred; XXX034
\item parent -- reads to parent envs of the local one. XXX035
\item call-stack -- reads to envs that are on the call stack but not
  caller. XXX035.
\end{itemize}

In the corpus we observe \AllWritesRnd writes to variables of which
\EvalWritesRnd writes happen inside \eval. However, all writes are not
dangerous. Only writes to environments not local to the computation spawned by
\eval are side-effecting. These writes outlive the computation and hence are
visible outside it. The remaining writes are local to the computation. We
observe that \EvalSideEffectingWritesRnd writes inside evals are side-effecting.
This is only \EvalSideEffectingWritesEvalPerc of all variable writes inside
\eval and \EvalSideEffectingWritesAllPerc of all variable writes in the corpus.

An \eval is considered side-effecting if it performs a side-effecting write to a
variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
packages are side-effecting.

\subsection{Provenance.}
% Explain that some generate expression objects and other language objects?
There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
    \item \c{substitute} synthesizes ASTs from expressions by replacing
    symbols with their bindings in the specified environment.
    \item \c{match.call} captures a call and its arguments as expressions.
    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression. % Aldd \c{call} ?
    \item \c{expression}  creates a vector of expression
    objects from text.
    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
    \item \c{comp} compiles R code into R bytecode.
\end{itemize}
We show the distribution of the different provenances in Table~\ref{tab:provenance}. Note that one site could be passed arguments from different provenances depending on the invokation. For some arguments, we could not determine their provenance or it was a value or a symbol, for which the provenance is less significant. % Not clear... Or maybe add symbols in that table?
% One can create a symbol with as.name, as.symbol etc. That would be the provenance (but a symbol can also come from quote for instance...)

\begin{table}[h]
    \begin{tabular}{|c|c|c|}
        \hline
        & \#sites & \%sites \\
        \hline
        \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
        \hline
        \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
        \hline
        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\
        \hline
        \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
        \hline
        \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
        \hline
        \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
        \hline \hline
        Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\
        \hline
    \end{tabular}
    \caption{Provenances of \eval expressions. } \label{tab:provenance}
\end{table}

>>>>

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically aren't,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
\eval. For the other cases, either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps, in fact the median is only
\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
mostly irrelevant (unless a built-in operator is redefined).

\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments, as few as \packageMinimizedpercentparentframesc of these
expressions are evaluated in a function environment.

\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
occurs in 6\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations. This is typically used in a function environment,
\packageMinimizedpercentparentframese of the time to be precise.

@@

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls.Assignments are excluded from this form. They
occurs in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
of these expressions run in function environments.

@@

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statements
code blocks, which occur in only 4\% of sites. These are larger expressions, we
do not inspect contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
returns a dataframe that is the result of fitting the model described in a given
formula. This form subsumes \c{F(F(X))}, \c{FUN} and assignments. It is the
single most popular function invoked from \eval, it occurs in 7\% of the sites.
Each call does quite a lot of work with a 2K instructions median.

@@

The environment in which evaluation happens determines what is visible to the
computation started by \eval, and the potential reach of its side-effects. The
second argument to \eval is used to specify that environment. We propose the
following high-level classification:

\begin{itemize}
\item \emph{Function:} these environments correspond to some function active on
  the call stack and are obtained by functions such as \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} these are environments that built from data structures
  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
  \c{list2env} or \c{as.environment}.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environments of a loaded library.
\end{itemize}

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
function environment, with \c{global} as a distant second. This means that most
variable lookups and most side effects will take place amongst the local
variables of some function. But which function? Table~3 shows the offset from
the function that directly called \eval, thus 0 is the direct caller and 1 is
its parent. The data suggests that over 80\% of function environments are the
direct caller -- this means, from a user's point of view -- the environment of
the function where \eval occurs textually. It is interesting that some 1.5\% of
sites evaluate a code in an environment that is three frame or above from the
caller of \eval. This distance makes it more tricky to reason about the behavior
of code. As the actions of \eval may happen at a distance.

\begin{table}[h]\centering\small
\begin{minipage}{4cm}
  \begin{tabular}{r|r|r}\hline
 Environment & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
 Synthetic & ? & ? \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Environments per site} \label{tab:highlevelenvironments}
\end{minipage} \hspace{1cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{r|r|r}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{fn}
\end{minipage}\hspace{1cm}
\begin{minipage}{3.8cm}
\begin{tabular}{r|r|r} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}
\end{table}

The use of the global environment is likely split between intentional and
accidental use. Direct references using \c{globalenv()} or \c{.GlobalEnv} is
rare, only in \packageNbExplicitGlobalSites sites. We summarize that the
majority of uses of the global environment come from the fact that most of the
code in our corpus consists of scripts that are loaded at the top level, thus
any access of a caller frame may end up executing in the global environment. The
reason we make this point is that values that are stored in the global
environment are visible to all functions and are not reclaimed by the garbage
collector. So accidental uses could result in pollution of that name space.

Synthetic environments must be given a parent. The parent is used to lookup
variables that are not found the new environment (side effects go to the new
environment and not to its parent). Table 5 shows the parents of synthetic
environments. Over half are functions, the second most frequent case is packages.
??? Say more ????  FIX: so that we get to 100\%

%% The \c{imchange} function of package \c{imager} makes it possible to modify
%% images using a dedicated formula syntax.\footnote{Inspired by {map} in
%% package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
%% new environment that inherits from \c{parent.frame()} by default (classified
%% as 1+).
%% \begin{lstlisting}
%%  newenv <- new.env()
%%  ...
%%  fo <- parse(text=as.character(fo)[2])
%%  im[where] <- eval(fo,envir=newenv,enclos=env)
%% \end{lstlisting}

%% \c{adjCoef} in package \emph{actuar} find the root of an equation defined by
%% a function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
%% transformed into an auxiliary function \c{h2} that can be optimized. Here,
%% the list used for \c{envir} ensures the correspondance between the textual
%% arguments of \c{h} and the arguments of \c{h2}.
%% \begin{lstlisting}
%%  sh <- substitute(h)
%%  fcall <- paste(sh, "(x, y)")
%%  ...
%%  h2 <- function(x, y)
%%  eval(parse(text = fcall),
%%  envir = list(x = x, y = y),
%%  enclos = parent.frame(2))
%% \end{lstlisting}

\subsection{The origins of \eval}

There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
\item \c{substitute} synthesizes ASTs from expressions by replacing symbols
    with their bindings in the specified environment.
\item \c{match.call} captures a call and its arguments as expressions.
\item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
\item \c{expression} creates a vector of expression objects from text.
\item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
   expressions.
\end{itemize}
We show the distribution of the different provenances in
Table~\ref{tab:provenance}. Note that one site could be passed arguments from
different provenances depending on the invokation. For some arguments, we could
not determine their provenance or it was a value or a symbol, for which the
provenance is less significant.

\begin{table}[h]\small
\begin{tabular}{r|r|r} \hline
Origin  & \#sites & \%sites \\\hline
\c{substitute} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
\c{parse} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\ \hline
\c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
\c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\\hline
\c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\\hline
Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\\hline
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{table}
